Associations of environmental and human factors with altered rodent populations in West Africa: A scoping review

### 2021-03-26 reports/manuscript.Rmd
finishing cleaning pathogen

@@

Designations of species change over time as new evidence emerges. As included studies span 45 years disambiguation of historical species names was performed using Taxize...

@@

  + Aims
  + Published/field

@@

### Trap setup

Thirty-three studies did not describe the setup of traps. Twenty five studies placed transect lines of traps through study sites with thirteen studies placing traps along a line but not clearly describing a transect across a defined site. Sixteen studies purposefully sample houses, with four further studies purposefully placing traps in targeted locationes. Nine studies placed traps in a grid across single or multiple habitat types. Three studies described a stratified or randomised approach to placing traps. Trap setup was typically reported for ecology studies (80%) with transect lines the most commonly used approach. Trap setup was reported in 53% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (20%).

@@

Trapping effort was variably reported in included studies. Thirty studies did not report trapping effort. Thirty-six studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures. Thirty-five studies reported incomplete data on trapping effort (i.e. number of study nights or total number of trap-nights). Trapping effort was more completely reported from ecology studies (53% complete information and 20% no information) compared to zoonoses risk studies (15% complete information and 41% no information).

  + Calculating where possible for site/study effort

  + Habitat classification

@@

Across all West African study sites there were 59,343 trapped small mammals (592 were trapped outside of West Africa), 2,830 (4.8%) trapped individuals were identified to order level (Rodentia), 5,329 (9%) were identified to genus level, the remaining 51,184 (86%) were identified to species level. Of the 131 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (102) with Muridae (83) being the largest family of rodents, followed by Gliridae (6), Nesomyidae (4), Sciuridae (3), Ctenodactylidae (2) and Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (all 1). The remaining 29 species came from the orders of the Soricomorpha (25), Erinaceomorpha (3) and a single Afrosoricida.

More individuals were trapped in studies conducted to investigate the risk of zoonoses (35,892, 60%) than from ecology studies (23,451 40%). Animals trapped as part of ecological research were more often identified to species level (93%) than those trapped in zoonoses research (82%). The most commonly trapped genera of rodents were *Mastomys sp.* (22,665, 39.7%), *Rattus sp.* (7,718, 13.5%), *Mus sp.* (7,317, 12.8%), *Praomys sp.* (4,887, 8.6%) and *Arvicanthis sp.* (4,408, 7.72%). At the species level *Mastomys natalensis* (10,419, 20.2%) and *Mastomys erythroleucus* (7,184, 13.9%) were the most commonly trapped rodents, *Rattus rattus* (6,229, 12.1%), *Mus musculus* (6,030, 11.8%), *Arvicanthis niloticus* (4,101, 8%) and *Mastomys huberti* (3,106, 6%) made up the next most commonly trapped species. Presence and absence maps for these six most commonly trapped species are shown in Figure \@ref(fig:dist).

>>>>

The brand or description of the rodent trap was obtained from each study. For those using multiple devices all types were recorded. Type of trapping method. Effort of trapping

### Habitat classification

@@

Classification of species is variable over time as morphological and genetic information emerges. To handle species classification synonyms and reclassification the reported names of trapped species was compared to the Global Biodiversity Information Facility []. For species with updated taxonomies the synonym was converted to the accepted species name for all subsequent analyses. Where a subspecies is reported by study authors the species complex it belongs to will be used for subsequent analysis.

### Pathogen testing

@@

### Aims of included studies

The detailed aims of the included studies are shown following division into the themes of ecological or zoonosis research. Biodiversity of rodents was a shared aim between the two themes with 37 ecological studies and 2 zoonoses studies investigating this. The remaining aims were more disparate with ecological studies investigating rodent population dynamics, rodent species assemblages, the genetics of rodents and their behaviours. Most zoonotic disease research focussed on prevalence of pathogens in rodent populations, the risk of spillover to human populations, the identity of pathogen reservoir species with fewer studies investigating the phylogenetics of pathogen strains (see Figure \@ref(fig:aims).

``` {r aims, fig.cap = "The aims of included studies following division into ecological and zoonotic infectious disease themes. Multiple aims were identified from each study, the number on the arrow indicates the number of distinct studies with that aim ", echo = F, layout = "l-page"}
knitr::include_graphics(here::here("figures", "study_aims.png"))
```

@@

### Trap setup within the study area

Thirty-three studies did not describe the structure of the trap setup within the defined study area. Twenty five studies placed transect lines of traps through study sites with thirteen studies placing traps along a line but not clearly describing a transect across a defined site. Sixteen studies purposefully sampled houses, with four further studies purposefully placing traps in targeted locations (i.e. near rivers, in fields, by houses). Nine studies placed traps in a structured grid across a single or multiple specified habitat types. Three studies described a stratified or randomised approach to placing traps within a defined area. Trap setup was commonly reported in ecology studies (80%) with transect lines the most commonly used approach. Trap setup was reported in 53% of zoonoses risk studies with purposeful placement of traps in houses the most common approach (20%).

@@

Trapping effort was variably reported across included studies. Thirty studies did not report any measures of trapping effort. Thirty-four studies reported a measure of trapping effort at the same level of detail at which they reported rodent captures  (the unit of analysis) and were classified as complete reports. Thirty-seven studies had incomplete reports on trapping effort due to not presenting the number of trap nights at the appropriate unit of analysis (i.e. number of study nights only or total number of trap-nights). Trapping effort was more completely reported from ecology studies (51% complete information and 20% no information) compared to zoonoses risk studies (13% complete information and 41% no information).

Among studies that reported complete trap-night information the median number of trap-nights per study was 4,764 (Interquartile range (IQR) = 1,227-10,477) with a minimum of 240 trap-nights and maximum of 45,274 trap-nights. For studies reporting the number of trap-nights at the unit of analysis (typically study site or habitat type) the median number of trap-nights was 360 (IQR = 246-793). In studies with incomplete reports on trap-nights at the unit of analysis there was a median of 7,893 trap-nights (IQR = 1,401-10,331) conducted, for 21 studies reporting only on the number of days or nights trapping was performed on activities typically lasted between 1-4 nights.

### Habitat classification

No studies classified habitats with reference to a specific habitat classification scheme. Seven studies did not describe the local habitat where traps were placed (3 ecology, 4 zoonoses). The remaining studies used greater than 160 descriptions of the habitat type in which trapping was conducted, further, the resolution of the description ranged from the habitat of the location of the trap through to the study site. The 160 reported habitat types were combined into 32 categories (see Supplementary table 2 for the habitat dictionary) for further analysis. At least one habitat was recorded for 16,168 trap sites (95%), with two or more habitats for a single trap site recorded for 4,195 (26%) sites. As single sites could span multiple habitat types the number of habitat type trap site combinations was 20,369. The most commonly trapped sites were in or around buildings (31%), in areas described as the rodents "natural habitat" (26%), in agricultural areas (13%) (e.g. rice fields, palm plantations), forests (6%) and within or around villages (5%).

Habitat type was more comprehensively reported from ecology than from zoonoses studies. Thirty-one habitat categories were reported from ecology studies with the most commonly investigated habitats being agriculture (23%), in or around buildings (18%) and within forests (15%). Among studies on zoonoses 15 habitat categories were reported, the most commonly reported classification was described as the rodents "natural habitat" (39%), within or around buildings (38%) and in agricultural settings (7%).

@@

Across all West African study sites there were 59,343 trapped small mammals (592 were trapped outside of West African countries), 2,830 (4.8%) trapped individuals were identified to order level (Rodentia), 5,329 (9%) were identified to genus level, the remaining 51,184 (86%) were identified to species level. Of the 124 distinct identified species trapped (see Supplementary table 1) the majority were from the order Rodentia (96) with Muridae (79) being the largest family of rodents, followed by Gliridae and Nesomyidae (both 4), Sciuridae (3), Ctenodactylidae (2) and Anomaluridae, Dipodidae, Hystricidae and Thryonomidae (all 1). The remaining 29 species came from the orders of the Soricomorpha (25), Erinaceomorpha (2) and Afrosoricida (1).

More individuals were trapped in studies conducted to investigate the risk of zoonoses (35,892, 60%) than from ecology studies (23,451 40%). Animals trapped as part of ecological research were more often identified to species level (93%) than those trapped in zoonoses research (82%). The most commonly trapped genera of rodents were *Mastomys sp.* (22,665, 39.7%), *Rattus sp.* (7,718, 13.5%), *Mus sp.* (7,324, 12.8%), *Praomys sp.* (4,977, 8.7%) and *Arvicanthis sp.* (4,408, 7.72%). At the species level *Mastomys natalensis* (10,351, 20.2%) and *Mastomys erythroleucus* (7,052, 13.8%) were the most commonly trapped rodents, *Rattus rattus* (6,228, 12.2%), *Mus musculus* (6,030, 11.8%), *Arvicanthis niloticus* (4,086, 8%) and *Mastomys huberti* (3,106, 6.1%) made up the next most commonly trapped species. Presence and absence maps for these six most commonly trapped species are shown in Figure \@ref(fig:dist).
